Detecting Stylistic Variation in Pop Production

### 2023-11-17 outputs/paper/paper.qmd
line edits, fixed first half of plots and tables

---
title: "Detecting Stylistic Variation in Pop Production using the Spotify API"
author: 
  - Michaela Drouillard
thanks: "Code and data are available at: https://github.com/michaeladrouillard/spotify_pop."
date: today
abstract: "This paper explores whether or not we can predict whether a song was produced by Jack Antonoff using the audio features provided by Spotify's API. We find that in general our logistic regression model has an accuracy score of 76%, a precision score of 39%, and a recall of 75.61%. Our random forest has an accuracy of 61%, and a No Information Rate of 43%. We interestingly found that danceability, a metric trained on human-tagged data, was a key predictor. Our results have implications for computational explorations of culture, how the combination human subjectivity and machine listening can come to create an approximate for style."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

We're interested in understanding how to create an approximate for production style. To do this we used Spotify API's audio features dataset to explore different audio feature metrics and identify which were the most import predictors in identifying pop producers' style. We chose Jack Antonoff as focus, as he is a prolific and widely used producer, who also has his own solo acts.

@@

We are interested in whether Spotify audio features can be used to train a model that distinguishes whether a song was produced by Antonoff. We obtained tracks from six artists who've collaborated with Antonoff, both before their collaboration and during. We also obtained tracks from similarly prolific pop producers. The producers we compare Antonoff's production style with are Little, Rechstaid, Martin, Kurstin, Epworth and Nowels. Our data set contains 3738 tracks in total.

We built two models to explore the data. The first model, a logistic regression, draws on the dataset of artists who've collaborated with Antonoff, with a binary variable to tag whether or not the song was an Antonoff production. With this model, we explore which variables are stronger predictors of Antonoff's style, developping a better understanding of whether the Spotify data can trace Antonoff's influence on an artist's sound over time. We built a second model, a Random Forest (RF), to complement the insights drawn from the logistic regression. The RF trains on the full data set, including other major pop producers.

@@

Against the odds, we can capture something about Antonoff's music with these variables, and it might, in the end, just be that subjective "ick" that some people feel, and others enjoy, and that we're all just trying to describe in using different methods and tools. The remainder of this paper is structured as follows:

@ref-data

@ref-Model

@ref-Results

@ref-Discussion

@ref-References

@@

We collected data from the Spotify API, which we accessed using the `spotifyr` package in R [@citeSpotify]. We used the `get_artist_audio_features` function to acquire data on Lorde, Taylor Swift, St. Vincent, Lana Del Rey, The Chicks, Florence and the Machine, and Bleachers discographies. HAIM, Marina and the Diamonds, Maggie Rogers, Sharon Von Etten, and Mitski were included to represent other contemporary pop artists with overlapping fan bases who have never collaborated with Antonoff. Of the artists who did collaborate with Antonoff, we chose artist who have produced multiple albums, and at least one album with Antonoff.

@@

tbl |>
  kable(col.names = c("Producer", "Number of Tracks"))

@@

The `get_artist_audio_features` provides a profile of each song, according to variables developped by Spotify, which makes for simple comparisons across artists. The final audio features in this study data set contains the variables: `artist_name`, `track_name`, `energy`, `danceability`, `key`, `loudness`, `mode`, `speechiness`, `acousticness`, `instrumentalness`, `liveness`, `valence`, `tempo`, `jack`.

@@

lr_data |>
  group_by(is_antonoff) |>
  summarise(across(-c(tempo, loudness), mean)) |>
  pivot_longer(cols = -is_antonoff) |>
  ggplot(aes(value, name, fill = factor(is_antonoff))) +
  geom_col(alpha = 0.8, position = "dodge") +
  theme_minimal() +
  labs(fill = "Jack", x = "Mean Value", y = "Feature")
```

@@

```{r}
#| label: fig-cdf
#| fig-cap: Cumulative Distribution Function of Each Variable, 
#| echo: false
#| warning: false
av<- read_csv(here::here("inputs/data/antonoff_viz.csv"))

@@

```{r}
#| label: fig-energy
#| fig-cap: Energy Variable
#| echo: false
#| warning: false

@@

```{r}
#| label: fig-danceability
#| fig-cap: Danceability Variable
#| echo: false
#| warning: false

@@

```{r}
#| label: fig-valence
#| fig-cap: Valence Variable
#| echo: false
#| warning: false

@@

```{r}
#| label: fig-acousticness
#| fig-cap: Acousticness Variable
#| echo: false
#| warning: false

@@

```{r}
#| label: fig-instrumentalness
#| fig-cap: Instrumentalness Variable
#| echo: false
#| warning: false

@@

```{r}
#| label: fig-tempo
#| fig-cap: Tempo Variable
#| echo: false
#| warning: false

@@

The aim of the logistic regression is to determine whether we can predict Antonoff's influence in artists' sound. We select a total of 1465 tracks from the dataset from 12 artists: Lorde, Taylor Swift, St. Vincent, Lana Del Rey, The Chicks, Florence and the Machine, HAIM, Marina and the Diamonds, Maggie Rogers, Sharon Von Etten, Mitski, and Bleachers (Antonoff's own band)..

@@

>>>>

---
title: "Detecting Stylistic Variation in Pop Production"
author: 
  - Michaela Drouillard
thanks: "Code and data are available at: https://github.com/michaeladrouillard/spotify_pop."
date: today
abstract: "We explore whether we can predict if a song was produced by Jack Antonoff using the audio features provided by Spotify's API. Our logistic regression model has an accuracy score of 76%, a precision score of 39%, and a recall of 76%. Our random forestmodel has an accuracy of 61%, and a No Information Rate of 43%. We especially found that danceability, a metric trained on human-tagged data, was a key predictor. Our results have implications for computational explorations of culture, how the combination of human subjectivity and machine listening can come to create an approximate for style."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

We are interested in understanding how to create an approximate for production style. To do this we used Spotify API's audio features dataset to explore different audio feature metrics and identify which were the most import predictors in identifying pop producers' style. We focus on Jack Antonoff, as he is a prolific and widely used producer, who also has his own solo acts.

@@

We are interested in whether Spotify audio features can be used to train a model that distinguishes whether a song was produced by Antonoff. We obtained tracks from six artists who've collaborated with Antonoff, both before their collaboration and during. We also obtained tracks from similarly prolific pop producers: Joel Little, Ariel Rechstaid, Max Martin, Greg Kurstin, Paul Epworth and Rick Nowels. Our data set contains 3738 tracks in total.

We built two models to explore the data. The first model, a logistic regression, draws on the dataset of artists who have collaborated with Antonoff, with a binary variable to tag whether or not the song was an Antonoff production. With this model, we explore which variables are stronger predictors of Antonoff's style, developping a better understanding of whether the Spotify data can trace Antonoff's influence on an artist's sound over time. We built a second model, a Random Forest (RF), to complement the insights drawn from the logistic regression. The RF trains on the full data set, including other major pop producers.

@@

Our results suggest that we can capture something about Antonoff's music with these variables, and it might, in the end, just be that subjective "ick" that some people feel, and others enjoy, and that we're all just trying to describe in using different methods and tools. The remainder of this paper is structured as follows:

@@

We collected data from the Spotify API, which we accessed using the `spotifyr` package in R [@citeSpotify]. We acquired data on Lorde, Taylor Swift, St. Vincent, Lana Del Rey, The Chicks, Florence and the Machine, and Bleachers discographies. HAIM, Marina and the Diamonds, Maggie Rogers, Sharon Von Etten, and Mitski were included to represent other contemporary pop artists with overlapping fan bases who have never collaborated with Antonoff. Of the artists who did collaborate with Antonoff, we chose artist who have produced multiple albums, and at least one album with Antonoff.

@@

tbl |>
  kable(col.names = c("Producer", "Number of Tracks"),
  digits = 1,
  booktabs = TRUE,
  linesep = "")

@@

Spotify provides variables for each song, according to variables developed by Spotify, which makes for simple comparisons across artists. The final audio features in our data set are: `artist_name`, `track_name`, `energy`, `danceability`, `key`, `loudness`, `mode`, `speechiness`, `acousticness`, `instrumentalness`, `liveness`, `valence`, `tempo`, `jack`.

@@

lr_data %>%
  select(-producer, -tempo, -loudness) %>%  
  pivot_longer(cols = -is_antonoff, names_to = "feature", values_to = "value") %>%
  ggplot(aes(x = value, y = feature, color = factor(is_antonoff))) +
  geom_point(alpha = 0.8, position = position_jitter(width = 0.2)) +
  scale_color_manual(values = c("blue", "orange"), labels = c("No", "Yes")) +
  theme_minimal() +
  labs(color = "Antonoff", x = "Value", y = "Feature")

@@

```{r, fig.width= 8, fig.height=6}
#| label: fig-cdf
#| fig-cap: Cumulative Distribution Function of Each Variable 
#| echo: false
#| warning: false
av<- read_csv(here::here("inputs/data/antonoff_viz.csv"))

@@

```{r, fig.width= 8, fig.height=6}
#| label: fig-energy
#| fig-cap: Energy Variable
#| echo: false
#| warning: false

@@

```{r, fig.width= 8, fig.height=6}
#| label: fig-danceability
#| fig-cap: Danceability Variable
#| echo: false
#| warning: false

@@

```{r, fig.width= 8, fig.height=6}
#| label: fig-valence
#| fig-cap: Valence Variable
#| echo: false
#| warning: false

@@

```{r, fig.width= 8, fig.height=6}
#| label: fig-acousticness
#| fig-cap: Acousticness Variable
#| echo: false
#| warning: false

@@

```{r, fig.width= 8, fig.height=6}
#| label: fig-instrumentalness
#| fig-cap: Instrumentalness Variable
#| echo: false
#| warning: false

@@

```{r, fig.width= 8, fig.height=6}
#| label: fig-tempo
#| fig-cap: Tempo Variable
#| echo: false
#| warning: false

@@

The aim of the logistic regression is to determine whether we can predict Antonoff's influence in artists' sound. We select a total of 1465 tracks from the dataset from 12 artists: Lorde, Taylor Swift, St. Vincent, Lana Del Rey, The Chicks, Florence and the Machine, HAIM, Marina and the Diamonds, Maggie Rogers, Sharon Von Etten, Mitski, and Bleachers (Antonoff's own band).

@@

```{r, fig.width= 8, fig.height=6}
#| echo: false
#| warning: false
